On probabilistic one-vs-one classification

### 2025-04-07 paper.tex
Clean up notation for coupling methods

Equations forming system \eqref{eq:bt2} are called Bradley-Terry equations.
The system is usually inconsistent and additional assumptions are needed. A method to solve Bradley-Terry equations is called a coupling method. Its input can be represented by a matrix $\boldsymbol{R}(x)$ with off-diagonal entries being $r_{ij}(x)$. Schematically, classification using coupling method $\boldsymbol{C}$ proceeds as indicated in Figure \ref{fig:coupling}.

\begin{figure}[!h]
	\centering
	\begin{tikzcd}[column sep=3cm]
		\textrm{sample $x$}\arrow[r, "binary~classifiers"] \arrow{dr}[swap]{multi-class~ prediction} & \boldsymbol{R}(x) \arrow{d}{coupling}[swap]{V} \\
		 & V(\boldsymbol{R}(x)) 
	\end{tikzcd}
	\caption{Multi-class classification using a coupling method ${V}$}
	\label{fig:coupling}
\end{figure}

@@

\emph{Symmetry} refers to the action of the permutation group which acts both on the classes as well as on multi-class predictions. The action of a permutation $\sigma$ extends to the action on the pairwise matrix $\boldsymbol{R} = (r_{ij})$ by 
\begin{equation}
	 r^\sigma_{i,j} \stackrel{def}{=}  r_{\sigma(i), \sigma(j)}.
\end{equation}
We say that a coupling method $V$ is \emph{symmetric} if 
\begin{align}
		V(\boldsymbol{R})^\sigma = V(\boldsymbol{R}^\sigma)\quad\textrm{for any permutation $\sigma$}.
\end{align}

@@

When we use a coupling method and the prior changes, we  can apply \eqref{eq:changePrior} either before or after coupling. If the results are identical, for all initial pairwise data $\boldsymbol{R}$, then we say that the coupling method is Bayes covariant. Symbolically, let us denote by  $\pi$ the change of priors. Let $\boldsymbol{R}^\pi$ be pairwise datum when we apply \eqref{eq:changePrior} to each binary classifier.
For Bayes covariant method $V$ the following diagram commutes.

\begin{figure}[!h]
\centering
\begin{tikzcd}
	\boldsymbol{R} \arrow[r, "\pi"] \arrow[d, "V"] & \boldsymbol{R}^\pi \arrow[d, "V"] \\
	V(\boldsymbol{R}) \arrow[r, "\pi"] & V(\boldsymbol{R}^\pi) = V(\boldsymbol{R})^\pi
\end{tikzcd}
\caption{Diagrammatical description of Bayes covariance for a coupling method $V$}
\label{fig:bc}
\end{figure}

@@

A natural way to find  $\hat{\boldsymbol{p}}$ is thus to minimize a functional incorporating divergence measures between these distributions
\begin{align*}
\hat{\boldsymbol{p}} = \argmin_{\boldsymbol{p}} \sum_{ij} w_{ij} \divm (\hat D_{ij}^\textrm{multi}, \hat D_{ij}^\textrm{binary})
\end{align*}

@@

For $K=3$, there are three different arboreal coupling methods, which we shall call $T_1$, $T_2$, $T_3$, where the system \eqref{eq:arb1} represents coupling method $T_2$.

@@

In order to convert Hinton's oracle into a true classification method we can use the principle of self-referentiality.  Denote by ${T}_i$ the prediction of the arboreal method, where the underlying tree is a star centered at class $C_i$. The predictions will generally span the vector space $R^K$. In that case the prediction $\hat{\boldsymbol{p}}$ of the method  can be expressed as a linear combination of $T_i(x)$. Let us now define \emph{radial} coupling method. The  method requires that for the  weights we could in fact use the components of $\boldsymbol{p}$ i.e.
\begin{equation}
	\begin{split}
	\hat {\boldsymbol{p}} &= \hat p_1 T_1(x) + \ldots + \hat p_K T_K(x)\\
	1 &= \hat p_1 + \ldots + \hat p_K
	\end{split}
	 \label{eq:radial}
\end{equation}
%
The intuition behind \eqref{eq:radial} is that if some component of $\hat{\boldsymbol{p}}$ is close to 1 (and thus the remaining components are close to zero), then the prediction $\hat{\boldsymbol{p}}$ should be close to the prediction of Hinton's oracle which is provided by ${{T}}_i$.

@@

There are many Bayes covariant methods. For instance any arboreal method is Bayes covariant.

@@

 Suppose real numbers $\alpha_1, \alpha_L$ are real numbers and $\boldsymbol{C}_1, \ldots, \boldsymbol{C}_L$ are coupling methods. We define their linear combination by requiring that 
 \begin{align*}
 	\bigl(\alpha_1 \boldsymbol{C}_1 \oplus \ldots \oplus \alpha_L \boldsymbol{C}_L\bigr)(x) \propto 
 \end{align*}

 given such $\alpha + \beta = 1$. If the prediction of the two Bayes covariant methods $\boldsymbol{B}$ and $\boldsymbol{\tilde B}$ are
\begin{align*}
	\boldsymbol{B}(x) = (p_1, \ldots, p_K)\quad\textrm{and}\quad \boldsymbol{\tilde B}(x) = (\tilde p_1, \ldots, \tilde p_K)
\end{align*}
then the prediction of the ensemble $\alpha 	\boldsymbol{B} \oplus \beta \boldsymbol{\tilde B}$ is given by
\begin{align*}
	\bigl(\alpha 	\boldsymbol{B} \oplus \beta \boldsymbol{\tilde B}\bigr)(x)\propto (\,p_1^\alpha \tilde p_1^\beta, \ldots, p_K^\alpha \tilde p_K^\beta).
\end{align*}

As an example, consider the classification problem with three classes. We have three arboreal classifiers $\boldsymbol{T}_1, \boldsymbol{T}_2, \boldsymbol{T}_3$ which correspond to star graphs centered respectively at classes $C_1, C_2, C_3$. The classifier 
\begin{align}
\frac13 \boldsymbol{T}_1 \oplus \frac13 \boldsymbol{T}_2 \oplus \frac 13 \boldsymbol{T}_3 = \frac13 \boldsymbol{T}_1 \oplus \frac23 \biggl(\frac12 \boldsymbol{T}_2 \oplus \frac 12 \boldsymbol{T}_3
\biggr), \label{eq:bc1}
\end{align}
is Bayes covariant \emph{and} symmetric. A key result proved in \cite{vsuch2016bayes} is that it is a unique such classifier for $K=3$.

@@

Let us point out one consequence of the theorem. We could define another ensemble $\boldsymbol{E}$ by averaging the predictions of $\boldsymbol{T}_1, \boldsymbol{T}_2, \boldsymbol{T}_3$:
\[
\boldsymbol{E} = \frac13 \boldsymbol{T}_1(x) + \frac13 \boldsymbol{T}_2(x) + \frac 13 \boldsymbol{T}_3(x).
\]
Note that here we use the plus symbol to denote the vector addition of vectors representing $K$-class probability distributions. This is a different classifier from the one given by  \eqref{eq:bc1} (see Appendix \ref{app:explicit}), and since it is symmetric, it cannot be Bayes covariant.

@@

It is natural to consider the following interrelated questions

\begin{itemize}
\item[a)] Which non-parametric coupling methods shows good performance across diverse data sets? Is it the orthodox method of Wu-Lin-Weng?
\item[b)] If the answer to a) is a coupling method which is not Bayes covariant, can we find a parametric method Bayes covariant method which  matches its performance?
\item[c)] If the answer to b) is negative, how serious is the impact of the lack of Bayes covariance in practice?
\end{itemize}

The following sections attempt to gather answers to these questions.

@@

\subsection{Binary classifiers}

@@

For training LDA we used MASS library in R.

@@

\subsection{Experiment 1: Accuracy of parameter-less coupling methods on benchmark datasets}

@@

\subsection{Experiment 2: Accuracy of Bayes covariant methods with parameters}

@@

As a second step, we consider whether the three arboreal methods $ \boldsymbol{S}_1,~ \boldsymbol{S}_2,~ \boldsymbol{S}_3$ (which do not have any parameters) could bridge the gap. The statistics are shown in Table \ref{tab:step2}.

@@

In the third step we define two parameter family of Bayes covariant coupling methods
\begin{align}
a_1 \boldsymbol{S}_1 + a_2 \boldsymbol{S}_2 + a_3 \boldsymbol{S}_3,\quad\textrm{where } a_1 + a_2 +a_3 = 1
\end{align}
Note that the parameter space of this family is not compact. In order to facilitate computations we opted to conduct a grid search on points in the simplex where the coefficients $a_i$ are nonnegative $a_i \geq 0$.

@@

%\subsection{Wu-Lin-Weng's method}

>>>>

Equations forming system \eqref{eq:bt2} are called Bradley-Terry equations.
The system is usually inconsistent and additional assumptions are needed. A method to solve Bradley-Terry equations is called a coupling method. Its input can be represented by a matrix $\boldsymbol{R}(x)$ with off-diagonal entries being $r_{ij}(x)$. Schematically, classification using coupling method $\boldsymbol{v}$ proceeds as indicated in Figure \ref{fig:coupling}.

\begin{figure}[!h]
	\centering
	\begin{tikzcd}[column sep=3cm]
		\textrm{sample $x$}\arrow[r, "binary~classifiers"] \arrow{dr}[swap]{multi-class~ prediction} & \boldsymbol{R}(x) %\arrow{d}{coupling}[swap]{\boldsymbol{v}} \\
		\arrow{d}{coupling} \\
		 & \boldsymbol{v}(\boldsymbol{R}(x)) 
	\end{tikzcd}
	\caption{Multi-class classification using a coupling method $\boldsymbol{v}$}
	\label{fig:coupling}
\end{figure}

@@

\emph{Symmetry} refers to the action of the permutation group which acts both on the classes as well as on multi-class predictions. The action of a permutation $\sigma$ extends to the action on the pairwise matrix $\boldsymbol{R} = (r_{ij})$ by 
\begin{equation}
	 r^\sigma_{i,j} \stackrel{def}{=}  r_{\sigma(i), \sigma(j)}.
\end{equation}
We say that a coupling method $V$ is \emph{symmetric} if 
\begin{align}
		\boldsymbol{v}(\boldsymbol{R})^\sigma = \boldsymbol{v}(\boldsymbol{R}^\sigma)\quad\textrm{for any permutation $\sigma$}.
\end{align}

@@

When we use a coupling method and the prior changes, we  can apply \eqref{eq:changePrior} either before or after coupling. If the results are identical, for all initial pairwise data $\boldsymbol{R}$, then we say that the coupling method is Bayes covariant. Symbolically, let us denote by  $\pi$ the change of priors. Let $\boldsymbol{R}^\pi$ be pairwise datum when we apply \eqref{eq:changePrior} to each binary classifier.
For Bayes covariant method $\boldsymbol{v}$ the following diagram commutes.

\begin{figure}[!h]
\centering
\begin{tikzcd}
	\boldsymbol{R} \arrow[r, "\pi"] \arrow[d, "coupling"] & \boldsymbol{R}^\pi \arrow[d, "coupling"] \\
	\boldsymbol{v}(\boldsymbol{R}) \arrow[r, "\pi"] & \boldsymbol{v}(\boldsymbol{R}^\pi) =\boldsymbol{v}(\boldsymbol{R})^\pi
\end{tikzcd}
\caption{Diagrammatical description of Bayes covariance for a coupling method $\boldsymbol{v}$}
\label{fig:bc}
\end{figure}

@@

A natural way to find  $\hat{\boldsymbol{p}}$ is thus to minimize a functional incorporating divergence measures between these distributions
\begin{align*}
\hat{\boldsymbol{p}} \stackrel{def}{=} \argmin_{\boldsymbol{p}} \sum_{ij} w_{ij} \divm (\hat D_{ij}^\textrm{multi}, \hat D_{ij}^\textrm{binary})
\end{align*}

@@

For $K=3$, there are three different arboreal coupling methods, which we shall call $\boldsymbol{s}_1$, $\boldsymbol{s}_2$, $\boldsymbol{s}_3$, where the system \eqref{eq:arb1} represents the coupling method $\boldsymbol{s}_2$.

@@

In order to convert Hinton's oracle into a true classification method we can use the principle of self-referentiality.  Denote by $\boldsymbol{s}_i$ the prediction of the arboreal method, where the underlying tree is a star centered at class $C_i$. The predictions will generally span the vector space $R^K$. In that case the prediction $\hat{\boldsymbol{p}}$ of the method  can be expressed as a linear combination of $\boldsymbol{s}_i(x)$. Let us now define \emph{radial} coupling method. The  method requires that for the  weights we could in fact use the components of $\boldsymbol{p}$ i.e.
\begin{equation}
	\begin{split}
	\hat {\boldsymbol{p}} &= \hat p_1 \boldsymbol{s}_1(x) + \ldots + \hat p_K \boldsymbol{s}_K(x)\\
	1 &= \hat p_1 + \ldots + \hat p_K
	\end{split}
	 \label{eq:radial}
\end{equation}
%
The intuition behind \eqref{eq:radial} is that if some component of $\hat{\boldsymbol{p}}$ is close to 1 (and thus the remaining components are close to zero), then the prediction $\hat{\boldsymbol{p}}$ should be close to the prediction of Hinton's oracle which is provided by ${{T}}_i$.

@@

There are many Bayes covariant methods. First examples are provided by arboreal coupling methods.

\begin{prop}
Any arboreal method is Bayes covariant.
\end{prop}

@@

Suppose real numbers $\alpha_1, \alpha_L$ are real numbers and $\boldsymbol{v}_1, \ldots, \boldsymbol{v}_L$ are coupling methods. We define their linear combination $\bigoplus_i a_i \boldsymbol{v}_i$ as the coupling method that for any $x$ yields the probability distribution on $L$ classes that satisfies
\begin{align*}
 	\bigl(a_1 \boldsymbol{v}_1 \oplus \ldots \oplus a_L \boldsymbol{v}_L\bigr)(x) \propto \boldsymbol{v}_1^{a_1}(x) \odot \cdots \odot \boldsymbol{v}_L^{a_L}(x),
\end{align*}
where the symbol $\odot$ denotes componentwise multiplication. With this definition we have the following result.

\begin{prop}
Suppose the coupling methods $\boldsymbol{v}_1, \ldots, \boldsymbol{v}_L$ are Bayes covariant. If $\sum a_i = 1$, then $\bigoplus_i a_i \boldsymbol{v}_i$ is also a Bayes covariant coupling method.
\end{prop}

% 
% given such $\alpha + \beta = 1$. If the prediction of the two Bayes covariant methods $\boldsymbol{B}$ and $\boldsymbol{\tilde B}$ are
%\begin{align*}
%	\boldsymbol{B}(x) = (p_1, \ldots, p_K)\quad\textrm{and}\quad \boldsymbol{\tilde B}(x) = (\tilde p_1, \ldots, \tilde p_K)
%\end{align*}
%then the prediction of the ensemble $\alpha 	\boldsymbol{B} \oplus \beta \boldsymbol{\tilde B}$ is given by
%\begin{align*}
%	\bigl(\alpha 	\boldsymbol{B} \oplus \beta \boldsymbol{\tilde B}\bigr)(x)\propto (\,p_1^\alpha \tilde p_1^\beta, \ldots, p_K^\alpha \tilde p_K^\beta).
%\end{align*}

As an example, consider the classification problem with three classes. We have three arboreal classifiers $\boldsymbol{s}_1, \boldsymbol{s}_2, \boldsymbol{s}_3$ which correspond to star graphs centered respectively at classes $C_1, C_2, C_3$. The classifier 
\begin{align}
\frac13 \boldsymbol{s}_1 \oplus \frac13 \boldsymbol{s}_2 \oplus \frac 13 \boldsymbol{s}_3  \label{eq:bc1}
\end{align}
is Bayes covariant \emph{and} symmetric. A key result proved in \cite{vsuch2016bayes} is that it is a unique such classifier for $K=3$.

@@

Let us point out one consequence of the theorem. We could define another ensemble $\boldsymbol{E}$ by averaging the predictions of $\boldsymbol{s}_1, \boldsymbol{T}_2, \boldsymbol{T}_3$:
\[
\boldsymbol{E} = \frac13 \boldsymbol{s}_1(x) + \frac13 \boldsymbol{s}_2(x) + \frac 13 \boldsymbol{s}_3(x).
\]
Note that here we use the plus symbol to denote the vector addition of vectors representing $K$-class probability distributions. This is a different classifier from the one given by  \eqref{eq:bc1} (see Appendix \ref{app:explicit}), and since it is symmetric, it cannot be Bayes covariant.

@@

The choice of a coupling method is an important task for practitioners. We aim to help by posing the following interrelated questions

\begin{itemize}
\item[a)] Which non-parametric coupling methods shows good performance across diverse data sets? Is it the orthodox method of Wu-Lin-Weng?
\item[b)] If the answer to a) is a coupling method which is not Bayes covariant, can we find a parametric method Bayes covariant method which  matches its performance?
\item[c)] If the answer to b) is negative, is it better to apply change of priors before or after coupling? Additionally, how serious is the impact of the lack of Bayes covariance in practice?
\end{itemize}

These questions will be put to test in Sections \ref{sec:exp1}, \ref{sec:exp2}, and \ref{sec:exp3}.

@@

\subsection{Classification methodology}

@@

For training LDA we used MASS library in R. We implemented coupling methods in C++ using Eigen library. As arguments to the coupling methods we use log-odds, i.e. instead of $r_{ij}$ they expect
$$
\tilde r_{ij} = \log \biggl(\frac{1}{r_{ji}} - 1\biggr).
$$
The resulting matrix $\tilde{\boldsymbol{R}}$ is skew symmetric (with the convention of having zeroes on the diagonal).

This parametrization allows for precise representation of extremely small probabilities which are sometimes produced by LDA. These small probabilities might be expected given that a large fraction of binary problems is linearly separable.

Small probabilities generally have very little effect on the implementation of Bayes covariant methods which generally operate using log-odds. However small probabilities do affect both the method of Wu-Lin-Weng as well sa radial method, because then Assumption \ref{ass:1} fails to hold.

@@

\subsection{Experiment 1: Accuracy of parameter-less coupling methods on benchmark datasets} \label{sec:exp1}

@@

\subsection{Experiment 2: Accuracy of Bayes covariant methods with parameters}
 \label{sec:exp2}

@@

As a second step, we consider whether the three arboreal methods $ \boldsymbol{s}_1,~ \boldsymbol{s}_2,~ \boldsymbol{s}_3$ (which do not have any parameters) could bridge the gap. The statistics are shown in Table \ref{tab:step2}.

@@

In the third step we define two parameter family of Bayes covariant coupling methods
\begin{align}
a_1 \boldsymbol{s}_1 + a_2 \boldsymbol{s}_2 + a_3 \boldsymbol{s}_3,\quad\textrm{where } a_1 + a_2 +a_3 = 1
\end{align}
Note that the parameter space of this family is not compact. In order to facilitate computations we opted to conduct a grid search on points in the simplex where the coefficients $a_i$ are nonnegative $a_i \geq 0$.

@@

\subsection{Experiment 3: Effect of change of priors}  \label{sec:exp3}
%\subsection{Wu-Lin-Weng's method}
